Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2021-09-19 vertical-thermal-review.Rmd
traits table change-- "physiological" to "biophysical"

The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions. Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$-limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].

@@

Leaf traits shape $T_{leaf}$, leaf metabolism, and its thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).

>>>>

The propensity for $T_{leaf}$ to be elevated above $T_{air}$ is not expected to be uniform across time and space, but rather to vary with micrometeorological conditions. Under drought conditions (hot and dry, with higher-than-average solar radiation), when $g_s$ is limiting, there will be a greater propensity for $T_{leaf} - T_{air}$ to be elevated in the canopy compared to the understory (Fig.`r fig_leaf_T`).
This is because sun leaves are exposed to higher irradiance and VPD, and therefore have steeper $g_s$-limitation than shade leaves [Fig. `r fig_leaf_T`e, @leighInfluenceLeafSize2017; @fauset_differences_2018]. 
In contrast, under hot and wet conditions, higher wind speeds and $g_b$ enable cooling in the upper canopy, whereas lower wind and $g_b$ in the understory allow much less evaporative cooling  [Fig. `r fig_leaf_T`b,d, @robertsStomatalBoundaryLayerConductances1990a; @martin_boundary_1999; @leighInfluenceLeafSize2017;  @songTreeSurfaceTemperature2020]. 
Thus, while understory and within-canopy shade leaves can remain cooler under lower radiation, their environment is not conducive to shedding excess heat, which in combination with their physiological propensities may result in above-optimal $T_{leaf}$ under slight heat or drought stress, or when exposed to higher levels of radiation, for example during sunflecks [@leighInfluenceLeafSize2017; @songTreeSurfaceTemperature2020; @schymanskiStomatalControlLeaf2013b; @zweifelMiddayStomatalClosure2002].
In addition to lower wind speeds (Fig. `r fig_NEON_vertical`), higher RH in the understory and inner canopy (Fig. `r fig_NEON_vertical`) would also inhibit cooling, as $\lambda E$ is stifled under high RH, resulting in  greater $T_{leaf}$-$T_{air}$ [Fig. `r fig_leaf_T`c, @tibbittsHumidityPlants1979b; @songTreeSurfaceTemperature2020;  @perezIncreasingHumidityThreatens2018; @konradLeafTemperatureIts2021; @dietzVerticalPatternsDuration2007].

@@

Leaf traits shape $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015;@michaletzEnergeticCarbonEconomic2016;  @zwienieckiHydraulicLimitationsImposed2004].
Leaf morphology adjusts to the previously experienced microenvironment during leaf formation in buds, and also during ongoing leaf expansion [@zwienieckiHydraulicLimitationsImposed2004], and the ratio of far red to red light and associated water demands determines the differentiation of traits of sun and shade leaves [@casasSunShadeLeaves2011; @zwienieckiHydraulicLimitationsImposed2004; @keenanGlobalLeafTrait2016]. 
This dichotomy in leaves is observed along the vertical gradient where canopy-leaves vary dramatically from canopy-interior and understory due to the differences in biophysical conditions (Fig. `r fig_NEON_vertical`, Supporting Information Figure `r SIfig_NEON_all`).
